05 - PubliliaFitness - Estimating lifetime fecundity and fitness

### 2023-08-26 scripts/Publilia-05-fitness.Rmd
init `Publilia-05-fitness.Rmd`

---
title: "Publilia_2019_CHAPTER-2C-fitness"
author: "micah fletcher"
date: "2/24/2022"
output: html_document
---

# Publilia 2019 Chapter - Script 2C: estimating fitness
This is meant to consolidate previous scripts and provide a readable and reproducible record of the analyses for the Publilia Chapter.

## Lifetime fitness using regression predictions
Custom functions for using survivorship regression predictions to estimate fitness
```{r}
ci_for_fitness <- function(data_in, survNom, survDen, 
                        tenure, treatment, vis, model){
  data <- eval(substitute(
    data_in %>%
      add_column(fit = predict(model, newdata = ., type = 'response'))
    ))
  ## grad the inverse link function
  ilink <- family(model)$linkinv
  ## add fit and se.fit on the **link** scale
  data_fit_link <- eval(substitute(
    data %>% add_column(
  setNames(as_tibble(predict(model, data, se.fit = TRUE)[1:2]),
           c('fit_link','se_link'))) %>%
## create the interval and backtransform
  mutate(fit_resp  = ilink(fit_link),
         conf_upr = ilink(fit_link + (2 * se_link)),
         conf_lwr = ilink(fit_link - (2 * se_link)))
    ))
  return(data_fit_link)
}
```

@@

Not time for dissertation but I could experiment with using a square root function for egg laying (more eggs in the first few days of guarding/visiting) and see how robust these results are to our assumptions about egg laying behavior.
```{r}
est_fitness_prep <- visited_1st_4d_data %>%
  filter(!is.na(initiation_date)) %>%
  group_by(Female_ID) %>%
  mutate(started_by_initiating =
           any(clutch_visited_order==1&clutch_initiated_order==1)) %>%
  ungroup() %>%
  select(location, Female_ID, clutch_visited_order, clutch_initiated_order, 
         desert_rel2h = est_rel2h, guard_duration,
         mom_at_init, started_by_initiating) %>%
  left_join(select(clutch_survivorship_data, location:last_desert, 
                   n_eggs, med_t5eggs, n_hatched, n_survived, DormPlant), 
            by=c("location")) %>%
  left_join(select(focalMb4h_sc2h, Female_ID, location, focalMb4h_sc2h),
            by=c("Female_ID", "location")) %>%
  mutate(des=last_des_rel2h, 
         ants=ifelse(ants_removed,"Removed","Control"),
         vis=`visited in 4 days`) %>%
  left_join(select(ant_counts, location, ever_antsP)) %>%
  separate(location, into = c("DormPlant", "Leaf"), sep="_", remove=FALSE) %>%
  left_join(dplyr::select(moms_plant_data, DormPlant, last_visit_ends_P), by="DormPlant") %>%
  mutate(last_desP_rel2h = last_visit_ends_P-est_hatch,
         ant_treatment=ifelse(ants_removed,"Removed","Control"))

@@

## Split violin plots
geom_split_violin
```{r}
GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin, 
                           draw_group = function(self, data, ..., draw_quantiles = NULL) {
  data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
  grp <- data[1, "group"]
  newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
  newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
  newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

  if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
    stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
      1))
    quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
    aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
    aesthetics$alpha <- rep(1, nrow(quantiles))
    both <- cbind(quantiles, aesthetics)
    quantile_grob <- GeomPath$draw_panel(both, ...)
    ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
  }
  else {
    ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
  }
})

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}
```

@@

plotting libraries
```{r}
library(multcompView)
```

geom_split_violin
```{r}
GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin, 
                           draw_group = function(self, data, ..., draw_quantiles = NULL) {
  data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
  grp <- data[1, "group"]
  newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
  newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
  newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

  if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
    stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
      1))
    quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
    aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
    aesthetics$alpha <- rep(1, nrow(quantiles))
    both <- cbind(quantiles, aesthetics)
    quantile_grob <- GeomPath$draw_panel(both, ...)
    ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
  }
  else {
    ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
  }
})

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}
```

A function to run Kruskall-Wallis test and Dunn tests for each panel
```{r}
fitness_KW_test <- function(data, y, group){
  # non-parametric: Kruskall-Wallis test followed by Dunn test
  # run KW test using rstatix::kruskal_test
  KWtest <- eval(substitute(data %>% kruskal_test(y ~ group)))
  # get effect size using effectsize::rank_epsilon_squared
  KWepsilon2 <- eval(substitute(rank_epsilon_squared(y ~ group, data=data)))
  # run Dunn test using rstatix::dunn_test
  dunn <- eval(substitute(data %>% dunn_test(y ~ group)))
  # get multiple comparison groups letters using multcompView::multcompLetters
  dunn.vec <- dunn$p.adj.signif
  names(dunn.vec) <- paste(dunn$group1,dunn$group2,sep="-")
  multcompLetters <- tibble(x=names(multcompLetters(dunn.vec)$Letters),
                            label=unname(multcompLetters(dunn.vec)$Letters))
  # return
  out <- list(KWtest=KWtest, KWepsilon2=KWepsilon2, dunn=dunn, multcompLetters=multcompLetters)
  return(out)
}

fitness_life_strats_data.itero <- filter(fitness_life_strats_data, n_clutches_with_eggs_iv>1)
fitness_future_stats_eggs.itero <- 
  fitness_KW_test(data=fitness_life_strats_data.itero, y=est_eggs_future, group=desert_cat_ants)

KWtest <- fitness_life_strats_data %>% kruskal_test(est_eggs_future ~ desert_cat_ants)

# # test
# fitness_life_strats_data <- est_fitness_life_strats %>% 
#   mutate(desert_cat_ants = ifelse(ants_removed, 
#                                   paste(desert_cat,"ants removed"),paste(desert_cat,"control")))
# fitness_KW_test(data=fitness_life_strats_data, y=est_eggs_life, group=desert_cat_ants)

# # Troubleshooting:
# fitness_life_strats_data %>% filter(!females_removed, !is.na(desert_cat)) %>% 
#     ggplot(data = ., aes(x = desert_cat, y = est_eggs_life, fill = ants_removed, 
#         color = ants_removed)) + ylim(-160 * 0.05, 160) + 
#   geom_split_violin(color = NA, alpha = 0.5, key_glyph = draw_key_rect) + 
#   geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.3), 
#              pch = 16, size = 2, key_glyph = draw_key_rect) + 
#     xlab("Guarding strategy for first clutch") + theme_publilia() + 
#     scale_fill_manual(labels = c("Control", "Ants removed"), 
#         values = c("brown1", "black")) + scale_color_manual(labels = c("Control", 
#     "Ants removed"), values = c("brown1", "black")) + 
#   geom_text(data = fitness_stats_eggs$multcompLetters, 
#             aes(x = x, y = -160 * 0.05, label = label), 
#             fontface = "bold", size = 4, position = position_dodge(width = 0.5)) + 
#   scale_x_discrete(labels = c("desert", "eggs", "nymphs", "visit"))
# 
# fitness_life_strats_data %>% filter(!females_removed, !is.na(desert_cat)) %>% 
#     ggplot(data = ., aes(x = desert_cat, y=est_eggs_life, fill=ants_removed, color=ants_removed)) + 
#   ylim(-160 * 0.05, 160) + 
#   geom_split_violin(color = NA, alpha = 0.5, key_glyph = draw_key_rect) + 
#   geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.3), 
#              pch = 16, size = 2, key_glyph = draw_key_rect) + 
#     xlab("Guarding strategy for first clutch") + theme_publilia() + 
#     scale_fill_manual(labels = c("Control", "Ants removed"), values = c("brown1", "black")) + 
#   scale_color_manual(labels = c("Control", "Ants removed"),  values = c("brown1", "black")) + 
#   geom_text(data = fitness_stats_eggs$multcompLetters, 
#     aes(x = x, y = -160 * 0.05, label = label), 
#     fontface = "bold", size = 4, position=position_dodge(width=0.5)) + 
#   scale_x_discrete(labels = c("desert", "eggs", "nymphs", "visit"))

# plot boxplots, including KWtest letters
fitness_plot_dunn <- function(data, y, group, itero=F, multcompLetters, letterSize, ymax=160){
  if(itero==F){
    (p <- eval(substitute(data %>%
      filter(!females_removed, !is.na(group)) %>%
    ggplot(data = ., aes(x = group, y=y, fill=ants_removed, color=ants_removed)) + 
      ylim(-ymax * 0.05, ymax) + 
      geom_split_violin(color = NA, alpha = 0.5, key_glyph = draw_key_rect) + 
      geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.3), 
                 pch = 16, size = 2, key_glyph = draw_key_rect) + 
      xlab("Guarding strategy for first clutch") + theme_publilia() + 
      scale_fill_manual(labels = c("Control", "Ant exclusion"), values = c("brown1", "black")) + 
      scale_color_manual(labels = c("Control", "Ant exclusion"),  values = c("brown1", "black")) +
      geom_text(data = multcompLetters, 
                aes(x = x, y = -ymax * 0.05, label = label), 
                fontface = "bold", size = letterSize, position=position_dodge(width=0.5)))))
    return(p)
  } else {
    # single-clutch females
    (p_semel <- eval(substitute(data %>%
      filter(!females_removed, !is.na(group),
             (n_clutches_with_eggs_iv==1)) %>%
        ggplot(data = ., aes(x = group, y=y, fill=ants_removed, color=ants_removed)) + 
          ylim(-ymax * 0.05, ymax) + 
          geom_split_violin(color = NA, alpha = 0.5, key_glyph = draw_key_rect) + 
          geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.3), 
                     pch = 16, size = 2, key_glyph = draw_key_rect) + 
          xlab("Guarding strategy for first clutch") + theme_publilia() + 
          scale_fill_manual(labels = c("Control", "Ant exclusion"), values = c("brown1", "black")) + 
          scale_color_manual(labels = c("Control", "Ant exclusion"),  values = c("brown1", "black")) +
          geom_text(data = multcompLetters, 
                    aes(x = x, y = -ymax * 0.05, label = label), 
                    fontface = "bold", size = letterSize, position=position_dodge(width=0.5)))))
    # multi-clutch females
     (p_itero <- eval(substitute(data %>%
      filter(!females_removed, !is.na(group),
             (n_clutches_with_eggs_iv>1)) %>%
        ggplot(data = ., aes(x = group, y=y, fill=ants_removed, color=ants_removed)) + 
          ylim(-ymax * 0.05, ymax) + 
          geom_split_violin(color = NA, alpha = 0.5, key_glyph = draw_key_rect) + 
          geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.3), 
                     pch = 16, size = 2, key_glyph = draw_key_rect) + 
          xlab("Guarding strategy for first clutch") + theme_publilia() + 
          scale_fill_manual(labels = c("Control", "Ant exclusion"), values = c("brown1", "black")) + 
          scale_color_manual(labels = c("Control", "Ant exclusion"),  values = c("brown1", "black")) +
          geom_text(data = multcompLetters, 
                    aes(x = x, y = -ymax * 0.05, label = label), 
                    fontface = "bold", size = letterSize, position=position_dodge(width=0.5)))))
    p_list<-list(semel=p_semel, itero=p_itero)
    return(p_list)
  }
}

# test
# fitness_life_strats_data <- est_fitness_life_strats %>%
#   mutate(desert_cat_ants = ifelse(ants_removed,
#                                   paste(desert_cat,"ants removed"),paste(desert_cat,"control")))
# eggs_fitness_stats <- fitness_KW_test(data=fitness_life_strats_data, y=est_eggs_life, group=desert_cat_ants)
# 
# fitness_plot_dunn(data=fitness_life_strats_data, y=est_eggs_life, group=desert_cat_ants,
#                   itero=F, multcompLetters=eggs_fitness_stats$multcompLetters,
#                   letterSize=8)
```

>>>>

---
title: '05 - PubliliaFitness - Estimating lifetime fecundity and fitness'
author: "Micah Fletcher"
date: '`r Sys.Date()`'
output: 
  rmdformats::robobook:
  fig_width: 9
toc_depth: 3
---

# Load libraries

```{r}
library(multcompView)
source(here("R/Publilia-05-fitness-functions.R"))
```

# 0 - Data import

```{r}
surv_reg_data <- read_rds(here("saved_rds/surv_reg_data.rds"))
```

@@

```{r}
est_fitness_prep <- visited_1st_4d_data %>%
  filter(!is.na(initiation_date)) %>%
  group_by(Female_ID) %>%
  mutate(started_by_initiating =
           any(clutch_visited_order==1&clutch_initiated_order==1)) %>%
  ungroup() %>%
  select(location, Female_ID, clutch_visited_order, clutch_initiated_order, 
         desert_rel2h = est_rel2h, guard_duration,
         mom_at_init, started_by_initiating) %>%
  left_join(select(clutch_survivorship_data, location:last_desert, 
                   n_eggs, med_t5eggs, n_hatched, n_survived, DormPlant), 
            by=c("location")) %>%
  left_join(select(focalMb4h_sc2h, Female_ID, location, focalMb4h_sc2h),
            by=c("Female_ID", "location")) %>%
  mutate(des=last_des_rel2h, 
         ants=ifelse(ants_removed,"Removed","Control"),
         vis=`visited in 4 days`) %>%
  left_join(select(ant_counts, location, ever_antsP)) %>%
  separate(location, into = c("DormPlant", "Leaf"), sep="_", remove=FALSE) %>%
  left_join(dplyr::select(moms_plant_data, DormPlant, last_visit_ends_P), by="DormPlant") %>%
  mutate(last_desP_rel2h = last_visit_ends_P-est_hatch,
         ant_treatment=ifelse(ants_removed,"Removed","Control"))

@@

@@
